Stability and Chaos in Multi-Agent Reinforcement Learning

### 2021-02-18 Documents/ChaoticDynamics/AAMAS 2021/Responses to Reviews for AAMAS.txt

We thank the reviewer for their comments. As for the questions:

We aim to focus our study on all normal form games. As you point out, whilst the strategy space of the agents includes a finite set of actions, they are able to play mixed strategies as well as pure ones. The goal of the paper is to identify the evolution of the strategy as agents learn. Thank you for pointing this out, we will clarify it in revision.

These are excellent points and we are, in fact, using the results of this study to consider how to 'steer' learning towards stable equilibria. As you mention, there are cases in which multiple stable equilibria exist. In such cases, we have not yet determined a closed form method for steering the system towards a given one. This is an important avenue for future work.

Finally, on the links with evolutionary dynamics (weaknesses), we notice that by deriving the continuous time dynamics of Q-Learning, [20] arrives at a system similar to the replicator dynamic. Our goal in this paper is primarily focused on understanding the stability of these dynamics.

@@

Review 2:

Thanks for the insightful remarks. Indeed, many of the techniques we apply are inspired by statistical physics. We believe that these methods can be fruitfully applied to learning on games (as discussed in Sec. 1), as the latter makes use of many random variables: strategies of agents, payoffs, etc. We aimed to focus the paper on the main, theoretical contributions, while providing details in the supplementary material. We thank you for taking the time to look at this material with such rigour.

@@

As regards the style of presentation, we are aware that normally, papers at AAMAS present results in a theorem-proof format. However, we felt that this format would not be appropriate for the results proved here, as the statement of each result would just amount to the last equation in the derivation.

Review 3:

We thank the reviewer for the several points they raise, they will be useful to improve the paper, as well as to provide some clarification.

The reviewer suggests there is a disconnect between the theory and experiments. 
Actually, the theoretical result (13) makes a number of predictions including: the decrease of stability for increasing alpha or tau, the (rather surprising) lack of dependence on Gamma (the correlation between payoffs), and that, for the limiting case of an infinite action set and a large number of players, the majority of the phase space is unstable. There are, however, practical restrictions when conducting experiments with a large action/player set. We do, however, present a number of empirical results that show the overall trend. Consider, e.g., Fig. 6 for 2-6 players, which shows the trend toward the predictions through increased instability. Therefore, (13) will hold in the limit, until then it overestimates the size of the unstable region, as in [15]. The message of the paper is to highlight the dependence of stability on parameters, and that algorithms must be analysed with stability in mind.

As regards cyclic behaviour in Prisoner's Dilemma, the study we refer to is actually different to that of Figure 1 and discusses a different learning algorithm (EWA) to the figure (Q-Learning). We will clarify this distinction.

@@

Method:

As discussed at the beginning of Sec. 3.1, i_mu (resp. i_kappa) refers to the action taken by player mu (resp. kappa). This allows us to forego using different letters for different players.

The point regarding scaling the system was given briefly on p. 4 to conserve space. A detailed discussion can be found in Section S2.2 of the supplementary material.

You are correct in saying that payoffs are already in a general form. To analyse the system we parameterise these payoffs by 'Gamma' which measures their correlation and study the behaviour of the average system which would result from a given Gamma (the effective dynamics (9)). This means that we can focus on how, on average, the payoff correlations affect stability rather than look at all the (infinite) possible variations of payoff matrices.

### 2021-02-18 Documents/ChaoticDynamics/AAMAS 2021/aamas21-formatting-instructions/main.tex
Three Body QL problems first added

\title[InstabilityinMARL]{Stability and Chaos in Multi-agent Reinforcement Learning}

@@

Single agent reinforcement learning (RL) is a well-established
framework for allowing agents to learn optimal strategies when trained
on an iterated task
%For a particularly strong example of this see
\cite{Vinyals2019}. However, for the realisation of complex tasks,
such as air traffic control, market negotiations, and multi-robot
coordination, it is required that the system be modelled as a
multi-agent system (MAS). Such systems are not as well-understood as
single-agent systems in RL, since a given agent is tasked with
optimising a reward function which depends not only on a
non-stationary environment, but also on the actions of other, possibly
loosely coupled, agents \cite{SchwartzMulti-agentApproach}.

It is, therefore, of paramount importance to develop a strong
theoretical understanding of multi-agent reinforcement learning (MARL)
to allow an {\em a priori} understanding of the behaviour of a given
learning algorithm. Fortunately, the study of MAS is not unique to the
field of AI and has been extensively investigated from the point of view of
economics and game theory as well
\cite{ShohamMultiagentFoundations}. In particular,
%the study of
evolutionary game theory (EGT) considers the problem of a MAS which is
repeatedly exposed to an iterated game. This idea shares a strong
resemblance with MARL and, in fact, in \cite{Tuyls2006AnGames} it was
shown that techniques from EGT may be fruitfully applied to the analysis of
Q-Learning \cite{Sutton2018}.

@@

    \item The agents have a large, but discrete, strategy space:
      during the theoretical study, we will make the assumption that
      the number $N$ of actions goes to infinity.

@@

Through our analysis and experiments we find that likelihood of
displaying unstable dynamics increases as the \textit{step length}
$\alpha$ and \textit{intensity of choice} $\tau$ parameters in the
Q-Learning algorithm
%\cite{Barber2012}
increase. However, the
correlation $\Gamma$ between payoffs, which measures how cooperative
or competitive a game is, does not affect the stability of the
system. In addition, we find that regardless of the choice of these
parameters, the likelihood of unstable behaviours increases as the
number  $p$ of players in the game increases.

@@

In this section we establish some of the preliminaries which
are required to follow the subsequent sections. The first regards
\textit{dynamical behaviours}, which considers the evolution of agent (note we use the terms \textit{agent} and \textit{player} interchangeably)
strategies as they learn how to play an iterated game. This study aims to
classify, based on the agent parameters and the payoff matrices, which
of these behaviours will be observed. The second is the \textit{dynamics
  of Q-Learning}, which are a set of equations that model the
aforementioned strategy evolution. It is on these dynamics that we
%\st{will}
perform our analysis.

@@

    As discussed in the Sec. \ref{sec::Intro}, when learning on iterated
    games, player strategies may exhibit much more complex behaviours
    than convergence to a Nash Equilibrium (NE), including convergence
    to a unique equilibrium (though not always to an NE), convergence
    to one of multiple equilibria, limit cycles, and chaos. These
    behaviours are illustrated in
    Fig.\ref{fig::DynamicalBehaviours}. To be able to predict the
    behaviour of a learning algorithm, it should ideally converge to a
    stable equilibrium although it is still possible to study systems
    with multiple equilibria or limit cycles
    \cite{Strogatz2000}. However, it would be difficult to control
    systems whose dynamics are governed by chaos (though research into
    controlling chaos is ongoing and rife with opportunity
    \cite{Fradkov2009}). It would, therefore, be a useful endeavour to
    determine the conditions under which the different sorts of behaviours
    arise.
%
    \begin{figure}[t]
    \centering
    \includegraphics[width = \linewidth]{Figures/DynamicalBehaviours.png}
    \caption{Different types of dynamical behaviour
       displayed
        by learning agents. Here the $x$-axis (resp.~$y$-axis) is the probability with which agent 1 (resp.~agent 2) chooses a given action. (Top Left) Figure drawn from \cite{Tuyls2006AnGames}.
        Convergence
        to a unique fixed point at the point (1, 1). (Top Right) Limit Cycle, the trajectories converge to cyclic behaviour (Bottom)
        Chaotic behaviour, here small deviations in the initial conditions can grow
        exponentially. (Top Right) and (Bottom) drawn from \cite{Sanders2018}.}
\end{figure}

@@

The behaviour of a system may be studied given a model of its
dynamics. It is through this process that a wide array of physical
systems, from harmonic pendulums to geophysical fluids, can be
understood. A growing body of research aims to understand multi-agent
reinforcement learning through the lens of its dynamics. In this
light, \cite{Tuyls2006AnGames} presents of derivation of a continuous-time dynamical system describing how agents following a Q-Learning
approach adjust the probabilities of choosing actions as they
iteratively play a game.

@@

\begin{equation}
    x_i(k) = \frac{e^{\tau Q_i(k)}}{\sum_j e^{\tau Q_j(k)}}
\end{equation}
%
where $\tau \in [0, \infty)$ is the \textit{intensity of choice} as described at the end of this section and $Q_i$ denotes the \textit{Q-value} of an action $i$, which is to be updated at each step according to
%
\begin{equation}
\label{eqn::Qupdate}
    Q_i(k+1) = (1 - \alpha) Q_i(k) + \alpha (r + \gamma \max_j Q_j(k))
\end{equation}
%
where $\alpha \in [0, 1]$ is the \textit{step length} parameter described below, $r$ is the immediate reward received, and $\gamma \in [0, 1]$ is the discount factor.

@@

Here,
agent 1 (resp.~agent 2) takes action $i$ with probability
$x_i$ (resp.~$y_j$).
When these actions are taken, the agents
receive payoff $a_{ij}$ and $b_{ji}$ respectively. With these equations, it is possible to
predict the expected behaviour of Q-Learning agents which they go on to empirically verify.

It is clear from (\ref{eqn::EOM}) that the long-term strategy
selection of these agents is determined by the parameters $\alpha,
\tau$ and the payoffs $a_{ij}, b_{ij}$. In light of this, this
contributions aims to address the question presented in the
introduction: how do these parameters influence the types of behaviours
seen during learning on an iterated game? Specifically, the parameters
we will consider are:
%
\begin{enumerate}
    \item $\alpha \in [0, 1]$: the \textit{step length}. Low values of $\alpha$ denote smaller
    updates. Heuristically, we can consider this to be the memory of the agent: lower $\alpha$
    denotes longer memory.

@@

    \item $\Gamma \in [-1, p-1]$: the \textit{payoff
      correlation}. Since there are an infinite number of realisations
      of the payoffs $a_{ij}, b_{ij}$, we instead analyse the
      behaviour for the average payoff case. To do this, we assume
      that the payoff matrices are drawn from a multi-variate Gaussian
      with mean zero and covariance matrix parameterised by
      $\Gamma$. We then average over this Gaussian. $\Gamma = -1$
      indicates a zero sum game, in which the sum of payoffs for a
      given action across all agents is 0, resulting in a purely
      competitive game. $\Gamma = p-1$ gives a purely cooperative game
      in which all agents share the same payoffs. The manner in which
      a game is generated from the choice of $\Gamma$ is described in
      (\ref{eqn::Payoffs}) below and follows the same procedure as
      outlined in \cite{Sanders2018}.
\end{enumerate}

@@

In this section we aim to determine how the choice of parameters
$\alpha$ and $\tau$, alongside the choice of payoff matrix, affect the
stability of the dynamics of Q-learning. However, as the values in the
payoff matrix can take any real number, there is an infinite number
of possible realisations of games. Of course, it would not be possible
to analyse every possible game, we instead follow the procedure
outlined in \cite{Coolen2005} and \cite{Galla2013} to average over
these realisations and, instead, analyse the \textit{effective
  dynamics}, the dynamics averaged over all realisations of payoff
matrices. Then, we perform a linear stability analysis around
equilibrium points to determine the conditions under which a fixed
point is stable. As this process is rather involved, we focus on the main steps in this paper, and refer to the Supplementary Material for the technical details. Sections from the Supplementary Material are denoted with the prefix S.

@@

\begin{enumerate}
    \item A trivial result is that the game is everywhere convergent (i.e. regardless of the choice of $\Gamma, N, p$) for the cases of $\alpha$ = 0 and/or $\tau$ = 0.  We see that choosing these values would result in the right hand side of (\ref{eqn::Final}) yielding a value of zero which falls within the stable region.
    \item Convergence is rare. This is seen by the fact that the analytic result overestimates the region of instability. In fact, for all allowed choices of $\Gamma, \alpha, \tau$, the right hand side of (\ref{eqn::Final}) is negative, which violates the stability criterion. This implies that games, in general, will not converge.
    \item The likelihood of convergence decreases with increasing $\alpha$ and $\tau$, regardless of the choice of $N, p$. We can see this since the right hand side of (\ref{eqn::Final}) tends further away from zero (i.e. further from the region of stability).
    \item The choice of $\Gamma$ does not affect the stability of the game in the negatively-correlated regime. Therefore, the only dependent factors are $\alpha, \tau, N, p$. This is more easily interpreted from the heatmaps in Figure \ref{fig:theory}.
    \item The likelihood of convergence decreases as $p$ increases. We see this by noticing the weighting that the $(p-1)$ in the second term of (\ref{eqn::Final}) places. As $p$ increases, this term drives the system further from the stability boundary.

@@

We illustrate these implications in Figure \ref{fig:theory} which plots the value obtained by the right hand side of (\ref{eqn::Final}). In order to determine these values, we first had to solve (\ref{eqn::fixed_point}) for $q$ and $\chi$. We determine the $\xfixed$ which solves the expression inside the square brackets of (\ref{eqn::fixed_point}) using a Newton-Raphson root-finding approach and use the following expressions for $q, \chi$ to determine their value. 
%
\begin{eqnarray*}
%    \begin{split}
        1 & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z) exp(-\frac{z^2}{2}) dz \\ 
        q & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z)^2 exp(-\frac{z^2}{2}) dz \\ 
        \chi & = & \frac{1}{q^{n/2}}\frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \frac{\partial \xfixed (z)}{\partial z} exp(-\frac{z^2}{2}) dz
%    \end{split}
\end{eqnarray*}

It should note that this method of finding $\xfixed$ yields only an approximate value and so the $q, \chi$ which are found are estimates of their true value.

We consider each of the implications (2)-(5) from the Sec. \ref{sec::Theory}
in turn, choosing to forego illustrating (1) as it can be seen immediately
from (\ref{eqn::Final}). $\tau$ is held fixed
at $0.05$ and vary $\Gamma$ in
the range $[-1, 0]$, $\alpha$ in the range $[0.01, 0.05]$.

\paragraph{Implication 2:} By looking at the values of the heatmap, we see that, for all choices
of the parameters
(\ref{eqn::Final}) evaluates to a negative value, which signals
instability. As such, the analysis over-estimates the region of
instability. We believe that this is due to the
%heavy
assumptions made during the derivation of the analytic result, the
strongest of which was to drop the discrepancy between players and
actions, treating each action of each player to be independent and
identically distributed. However, even within the error that this
assumption generates, it is still clear (and is verified
experimentally) that (\ref{eqn::Final}) predicts convergence in Q-Learning to be rare. Indeed as the
value of $N$ increases, the stability boundary determined by
experiment (Figure \ref{fig:NumericalExperiments}) tends towards
$\alpha = 0, \tau = 0$.

Similarly, as the value of $N$ increases, the values in the heatmap
(seen on the right of the map) reduce. This is due to the assumption
made in the derivation of the analytic result that $N \rightarrow
\infty$. As the value of N increases, the discrepancy between the true
behaviour of the system and analytic result decreases. This does,
however, give rise to a limitation of the result - namely that the
effect of the number of actions $N$ cannot be accurately inferred from
 (\ref{eqn::Final}). However, it should still be noted that, even
with the increase of $N$, the values in parameter space remain negative
and still, accurately, predict instability. We go on to verify that
the effect of $p, \alpha, \Gamma$ and $\tau$ are accurately captured
by (\ref{eqn::Final}).

We interpret (\ref{eqn::Final}), for given parameters, by determining
how close the right hand side is to the stability boundary (i.e., how
close the value is to zero). A choice of parameters which evaluates
close to the stability boundary has a higher chance (by the analytic
error) of being convergent than one which lies further from the
boundary. Therefore, the heat-maps in Figure \ref{fig:theory} should
be considered to show the probability of convergence for a choice of
$(\alpha, \Gamma)$. Lighter regions denote a higher probability of
convergence whereas darker regions are those whose values are further
from zero and therefore show a lower probability of convergence.

\paragraph{Implication 3} can be visualised by noticing that the likelihood of convergence decreases (the heatmap becomes darker) as $\alpha$ increases, and similarly for $\tau$. This occurs independent of the choice of $N$ or $p$.

\paragraph{Implication 4} can similarly be inferred by the trend of convergence, which occurs only in varying $\alpha$. The choice of $\Gamma$
makes almost no difference in the evaluation of (\ref{eqn::Final}). We
anticipate that this is due to the restriction of $\Gamma$ to be in
the range $[-1, 0]$. In this range, the scaling effect of $\Gamma^2$
is dominated by the $\alpha^4$ and $\tau^4$ terms which appear in
(\ref{eqn::Final}).

@@

\paragraph{Implication 5}, finally, is considered in Figure \ref{fig:pvariation}. In this we keep all parameters, except for $p$, fixed and evaluate the right hand side of (\ref{eqn::Final}) for varying $p$, taking the absolute value for convenience. We see that the trend is to tend away from the stability boundary, indicating that as the number of players in the system increases, so too does the region of instability.

@@

\begin{figure}[t]
    \includegraphics[width =  \linewidth, center]{Figures/alphavar5e2.png}
    \caption{The trajectories of action probabilities for 2 players with 5 actions trained on a game of $\Gamma = -0.5$ and $\tau = 0.05$. (Left) $\alpha = 0.01$, (Right) $\alpha = 0.05$.}
    \label{fig:AlphaVariation}
\end{figure}

@@

Finally, Figure \ref{fig:pVarianceExpt} plot the degree variation about the fixed point displayed as the number of players $p$ increases. The variation is determined by first allowing the game to iterate for 5000 steps, so that it can be assumed to have reached the vicinity of the fixed point. Then, the action probabilities are recorded for a further 5000 iterations. At the end of this second period, the variance of the actions are calculated as
%
\begin{equation*}
    V(t) = \frac{1}{N} \sum_i \frac{1}{5000} \sum_t x_i(t)^2 - \left[\frac{1}{5000} \sum_t x_i(t) \right]^2 .
\end{equation*}

@@

We note that the numerical experiments confirm the predictions
suggested by the analytic results in Figure
\ref{sec::Theory}. We notice that convergence occurs almost only for
low values of $\alpha$. This observation remains as we increase the
value of $N$. As noted, the analytic result over-estimates the region
of instability due to the assumptions of $N \rightarrow \infty$ and
$\frac{1}{p-1} << 2$ made during the derivation. Remaining
discrepancies are partially due to the aforementioned approximation in
the calculation of $q$, $\chi$ and the fact that the analytic result
considers a continuous time approximation of the expected behaviour of
Q-Learning. We expect that testing over a greater number of payoff
realisations and initial conditions will yield a more representative
assessment of the average behaviour of Q-Learning. However, running
these experiments is a computationally expensive procedure: for $p$
players and $N$ actions we require operations on $p$ matrices with
$N^{p}$ elements. As such, due to a reduced availability of
computational facilities, a large scale averaging was not possible.

>>>>

\title[InstabilityinMARL]{Stability and Chaos in Multi-Agent Reinforcement Learning}

@@

Single agent reinforcement learning (RL) is a well-established
framework for allowing agents to learn optimal strategies when trained
on an iterated task
%For a particularly strong example of this see
\cite{Vinyals2019}. For the realisation of complex tasks,
such as air traffic control, market negotiations, and multi-robot
coordination, however, it is required that the system be modelled as a
multi-agent system (MAS). Such systems are not as well-understood as
single-agent systems in RL, since a given agent is tasked with
optimising a reward function which depends not only on a
non-stationary environment, but also on the actions of other, possibly
loosely coupled, agents \cite{SchwartzMulti-agentApproach}.

It is, therefore, of paramount importance to develop a strong
theoretical understanding of multi-agent reinforcement learning (MARL)
to allow an {\em a priori} understanding of the behaviour of a given
learning algorithm. Fortunately, the study of MAS is not unique to the
field of AI and has been extensively investigated from the point of view of
economics and game theory as well
\cite{ShohamMultiagentFoundations}. In particular,
%the study of
evolutionary game theory (EGT) considers the problem of a MAS which is
repeatedly exposed to an iterated game. This idea shares a strong
resemblance with MARL and, in fact, in \cite{Tuyls2006AnGames} it was
shown that techniques from EGT may be fruitfully applied to the analysis of
Q-Learning \cite{Barber2012}.

@@

    \item The agents have a large, but discrete, strategy space:
      during the theoretical study, we will make the assumption that
      the number of actions $N$ goes to infinity.

@@

Through our analysis and experiments we find that likelihood of
displaying unstable dynamics increases as the \textit{step length},
$\alpha$ and \textit{intensity of choice}, $\tau$ parameters in the
Q-Learning algorithm
%\cite{Barber2012}
increase. However, the
correlation, $\Gamma$ between payoffs, which measures how cooperative
or competitive a game is, does not affect the stability of the
system. In addition, we find that regardless of the choice of these
parameters, the likelihood of unstable behaviours increases as the
number of players $p$ in the game increases.

@@

In this section, we establish some of the preliminaries which
are required in order to follow the subsequent sections. We first consider
\textit{dynamical behaviours}, which describes the evolution of agent (note we use the terms \textit{agent} and \textit{player} interchangeably)
strategies as they learn how to play an iterated game. This study aims to
classify, based on the agent parameters and the payoff matrices, which
of these behaviours will be observed. The second is the \textit{dynamics
  of Q-Learning}, which are a set of equations that model the
aforementioned strategy evolution. It is on these dynamics that we
%\st{will}
perform our analysis.

@@

    As discussed in the Sec. \ref{sec::Intro}, when learning on iterated
    games, player strategies may exhibit much more complex behaviours
    than convergence to a Nash Equilibrium (NE), including convergence
    to a unique equilibrium (though not always to an NE), convergence
    to one of multiple equilibria, limit cycles, and chaos. These
    behaviours are illustrated in
    Fig.\ref{fig::DynamicalBehaviours}. To be able to predict the
    behaviour of a learning algorithm, it should ideally converge to a
    stable equilibrium although it is still possible to study systems
    with multiple equilibria or limit cycles
    \cite{Strogatz2000}. It would be difficult, however, to control
    systems whose dynamics are governed by chaos (though research into
    controlling chaos is ongoing and rife with opportunity
    \cite{Fradkov2009}). It would, therefore, be a useful endeavour to
    determine the conditions under which the different sorts of behaviours
    arise.
%
    \begin{figure}[t]
    \centering
    \includegraphics[width = \linewidth, center]{Figures/DynamicalBehaviours.png}
    \caption{Different types of dynamical behaviour
       displayed
        by learning agents. Here the $x$-axis (resp.~$y$-axis) is the probability with which agent 1 (resp.~agent 2) chooses a given action. (Top Left) Figure drawn from \cite{Tuyls2006AnGames}.
        Convergence
        to a unique fixed point at the point (1, 1). (Top Right) Limit Cycle, the trajectories converge to cyclic behaviour (Bottom)
        Chaotic behaviour, here small deviations in the initial conditions can grow
        exponentially. (Top Right) and (Bottom) drawn from \cite{Sanders2018}. \label{fig::DynamicalBehaviours}}
\end{figure}

@@

The behaviour of a system may be studied given a model of its
dynamics. It is through this process that a wide array of physical
systems, from harmonic pendulums to geophysical fluids, can be
understood. A growing body of research aims to understand multi-agent
reinforcement learning through the lens of its dynamics. In this
light, Tuyls et al. \cite{Tuyls2006AnGames} present a derivation of a continuous-time dynamical system describing how agents following a Q-Learning
approach adjust the probabilities of choosing actions as they
iteratively play a game.

@@

\begin{equation}
    x_i(k) = \frac{e^{\tau Q_i(k)}}{\sum_j e^{\tau Q_j(k)}}
\end{equation}
%
where $\tau \in [0, \infty)$ is the \textit{intensity of choice} as described at the end of this section, and $Q_i$ denotes the \textit{Q-value} of an action $i$, which is to be updated at each step according to
%
\begin{equation}
\label{eqn::Qupdate}
    Q_i(k+1) = (1 - \alpha) Q_i(k) + \alpha (r + \gamma \max_j Q_j(k))
\end{equation}
%
where $\alpha \in [0, 1]$ is the \textit{step length} parameter described below, $r$ is the immediate reward received, and $\gamma \in [0, 1]$ is the discount factor.

@@

Here,
agent 1 (resp.~agent 2) takes action $i$ with probability
$x_i$ (resp.~$y_i$).
When these actions are taken, the agent
receives payoff $a_{ij}$ (resp.~$b_{ji}$). With these equations, it is possible to
predict the expected behaviour of Q-Learning agents, which they go on to empirically verify.

It is clear from (\ref{eqn::EOM}) that the long-term strategy
selection of these agents is determined by the parameters $\alpha,
\tau$ and the payoffs $a_{ij}, b_{ij}$. In light of this, this study
aims to address the question presented in the
introduction: how do these parameters influence the types of behaviours
seen during learning on an iterated game? Specifically, the parameters
we will consider are:
%
\begin{enumerate}
    \item $\alpha \in [0, 1]$: the \textit{step length}. Low values of $\alpha$ denote smaller
    updates. Heuristically, we can consider this to be the memory of the agent: lower $\alpha$
    denotes longer memory.

@@

    \item $\Gamma \in [-1, p-1]$: the \textit{payoff
      correlation}. Since there are an infinite number of realisations
      of the payoffs $a_{ij}, b_{ij}$, we instead analyse the
      behaviour for the average payoff case. To do this, we assume
      that the payoff matrices are drawn from a multi-variate Gaussian
      with mean zero and covariance matrix parameterised by
      $\Gamma$. We then average over this Gaussian. $\Gamma = -1$
      indicates a zero-sum game, in which the sum of payoffs for a
      given action across all agents is 0, resulting in a purely
      competitive game. $\Gamma = p-1$ gives a purely cooperative game
      in which all agents share the same payoffs. The manner in which
      a game is generated from the choice of $\Gamma$ is described in
      (\ref{eqn::Payoffs}) and follows the same procedure as
      outlined in \cite{Sanders2018}.
\end{enumerate}

@@

In this section, we aim to determine how the choice of parameters
$\alpha$ and $\tau$, alongside the choice of payoff matrix, affect the
stability of the dynamics of Q-learning. As the values in the
payoff matrix can take any real number, however, there is an infinite number
of possible realisations of games. Of course, it would not be possible
to analyse every possible game, so we instead follow the procedure
outlined in \cite{Coolen2005} and \cite{Galla2013} to average over
these realisations and, instead, analyse the \textit{effective
  dynamics}, the dynamics averaged over all realisations of payoff
matrices. Then, we perform a linear stability analysis around
equilibrium points to determine the conditions under which a fixed
point is stable. As this process is rather involved, we focus on the main steps in this paper, and refer to the Supplementary Material for the technical details. Sections from the Supplementary Material are denoted with the prefix S.

@@

\begin{enumerate}
    \item A trivial result is that the game is everywhere convergent (i.e. regardless of the choice of $\Gamma, N, p$) for the cases of $\alpha$ = 0 and/or $\tau$ = 0.  We see that choosing these values would result in the right hand side of (\ref{eqn::Final}) evaluating to zero, which falls within the stable region.
    \item Convergence is rare. This is seen by the fact that the analytic result overestimates the region of instability. In fact, for all allowed choices of $\Gamma, \alpha, \tau$, the right hand side of (\ref{eqn::Final}) is negative, which violates the stability criterion. This implies that games, in general, will not converge.
    \item The likelihood of convergence decreases with increasing $\alpha$ and $\tau$, regardless of the choice of $N$ and $p$. We can see this since the right hand side of (\ref{eqn::Final}) tends further away from zero (i.e. further from the region of stability).
    \item The choice of $\Gamma$ does not affect the stability of the game in the negatively-correlated regime. Therefore, the only dependent factors are $\alpha, \tau, N$ and $p$. This is more easily interpreted from the heatmaps in Figure \ref{fig:theory}.
    \item The likelihood of convergence decreases as $p$ increases. We see this by noticing the $(p-1)$ in the second term of (\ref{eqn::Final}). As $p$ increases, this term drives the system further from the stability boundary.

@@

We illustrate these implications in Figure \ref{fig:theory} which plots the value obtained by the right hand side of (\ref{eqn::Final}). In order to determine these values, we first had to solve (\ref{eqn::fixed_point}) for $q$ and $\chi$. We determine the $\xfixed$ which solves the expression inside the square brackets of (\ref{eqn::fixed_point}) using a Newton-Raphson root-finding approach and use the following expressions for $q$ and $\chi$ to determine their value. 
%
\begin{eqnarray*}
%    \begin{split}
        1 & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z) exp(-\frac{z^2}{2}) dz \\ 
        q & = & \frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \xfixed(z)^2 exp(-\frac{z^2}{2}) dz \\ 
        \chi & = & \frac{1}{q^{n/2}}\frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \frac{\partial \xfixed (z)}{\partial z} exp(-\frac{z^2}{2}) dz
%    \end{split}
\end{eqnarray*}

It should be noted that this method of finding $\xfixed$ yields only an approximate value and so the $q$ and $\chi$ which are found are estimates.

We consider each of the implications (2)-(5) 
in turn, choosing to forego illustrating (1) as it can be seen immediately
from (\ref{eqn::Final}). $\tau$ is held fixed
at $0.05$ and  $\Gamma$ is varied in
the range $[-1, 0]$ and $\alpha$ in the range $[0.01, 0.05]$.

\paragraph{Implication 2:} By looking at the values on the heatmap (Fig \ref{fig:theory}), we see that, for all choices
of the parameters,
(\ref{eqn::Final}) evaluates to a value less then zero, which signals
instability. As such, the analysis over-estimates the region of
instability. We believe that this is due to the
assumptions made during the derivation of the analytic result, the
strongest of which was to drop the discrepancy between players and
actions, treating each action of each player to be independent and
identically distributed. However, even within the error that this
assumption generates, it is still clear (and is verified
experimentally) that (\ref{eqn::Final}) predicts convergence in Q-Learning to be rare. Indeed, as the
value of $N$ increases, the stability boundary determined by
experiment (Figure \ref{fig:NumericalExperiments}) tends towards
$\alpha = 0, \tau = 0$.

Similarly, as the value of $N$ increases, the values in the heatmap
(seen on the right of the map) reduce. This is due to the assumption
made in the derivation of the analytic result that $N \rightarrow
\infty$. As the value of N increases, the discrepancy between the true
behaviour of the system and analytic result decreases. This does,
however, give rise to a limitation of the result - namely that the
effect of the number of actions, $N$, cannot be accurately inferred from
 (\ref{eqn::Final}). However, it should still be noted that, even
with the increase of $N$, the values in parameter space remain negative
and predict instability. We go on to verify that
the effect of $p, \alpha, \Gamma$ and $\tau$ are accurately captured
by (\ref{eqn::Final}).

We interpret (\ref{eqn::Final}), for given parameters, by determining
how close the right hand side is to the stability boundary (i.e. how
close the value is to zero). A choice of parameters which evaluates
close to the stability boundary has a higher chance (by the analytic
error) of being convergent than one which lies further from the
boundary. Therefore, the heatmaps in Figure \ref{fig:theory} should
be considered to show the probability of convergence for a choice of
$(\alpha, \Gamma)$. Lighter regions denote a higher probability of
convergence whereas darker regions are those whose values are further
from zero and therefore show a lower probability of convergence.

\paragraph{Implication 3:} can be visualised by noticing that the likelihood of convergence decreases (the heatmap becomes darker) as $\alpha$ increases, and similarly for $\tau$. This occurs independent of the choice of $N$ or $p$.

\paragraph{Implication 4:} can similarly be inferred by the trend of convergence, which occurs only in varying $\alpha$. The choice of $\Gamma$
makes almost no difference in the evaluation of (\ref{eqn::Final}). We
anticipate that this is due to the restriction of $\Gamma$ to
the range $[-1, 0]$. In this range, the scaling effect of $\Gamma^2$
is dominated by the $\alpha^4$ and $\tau^4$ terms which appear in
(\ref{eqn::Final}).

@@

\paragraph{Implication 5:} is considered in Figure \ref{fig:pvariation}. In this, we keep all parameters except for $p$ fixed, and evaluate the right hand side of (\ref{eqn::Final}) different values of $p$, taking the absolute value for convenience. We see that the trend is to tend away from the stability boundary indicating that, as the number of players in the system increases, so too does the region of instability.

@@

\begin{figure}[t]
    \includegraphics[width = 1.1 \linewidth, center]{Figures/alphavar5e2.png}
    \caption{The trajectories of action probabilities for 2 players with 5 actions trained on a game of $\Gamma = -0.5$ and $\tau = 0.05$. (Left) $\alpha = 0.01$, (Right) $\alpha = 0.05$.}
    \label{fig:AlphaVariation}
\end{figure}

@@

Finally, in Figure \ref{fig:pVarianceExpt} we plot the degree variation about the fixed point displayed as the number of players $p$ increases. The variation is determined by first allowing the game to iterate for 5000 steps, so that it can be assumed to have reached the vicinity of the fixed point. Then, the action probabilities are recorded for a further 5000 iterations. At the end of this second period, the variance of the actions is calculated as
%
\begin{equation*}
    V(t) = \frac{1}{N} \sum_i \frac{1}{5000} \sum_t x_i(t)^2 - \left[\frac{1}{5000} \sum_t x_i(t) \right]^2 .
\end{equation*}

@@

We note that the numerical experiments confirm the predictions
suggested by the analytic results in Figure
\ref{sec::Theory}. We notice that convergence occurs almost only for
low values of $\alpha$. This observation remains as we increase the
value of $N$. As noted, the analytic result over-estimates the region
of instability due to the assumptions of $N \rightarrow \infty$ and
$\frac{1}{p-1} << 2$ made during the derivation. Remaining
discrepancies are due to the aforementioned approximation in
the calculation of $q$, $\chi$ and the fact that the analytic result
considers a continuous time approximation of the expected behaviour of
Q-Learning. We expect that testing over a greater number of payoff
realisations and initial conditions will yield a more representative
assessment of the average behaviour of Q-Learning. However, running
these experiments is a computationally expensive procedure: for $p$
players and $N$ actions we require operations on $p$ matrices with
$N^{p}$ elements. As such, due to a reduced availability of
computational facilities, a large scale averaging was not possible.

However, due to the prevalence of non-convergent behaviour, shown by
the dark regions in Figure \ref{fig:NumericalExperiments} other than
at low values of $\alpha$, we conclude that implications (1),
(2) are verified. Indeed, implication (3) and (4) are also verified as
the proportion of convergent games decreases monotonically with
$\alpha$, but there does not appear to be any discernible
dependence on $\Gamma$. Finally, we can conclude that implication 5 is
verified by comparing Figures \ref{fig:pvariation} and
\ref{fig:pVarianceExpt}. It is clear that the instability of the
system increases monotonically with $p$. A point to note is that
Figure \ref{fig:pVarianceExpt} shows an almost logarithmic dependence of the
instability on $p$, whilst Figure \ref{fig:pvariation} suggests that
the dependence should be exponential. Due to the limited range of $p$
that was used to generate Figure \ref{fig:pVarianceExpt}, it cannot
yet be concluded that the dependence on $p$ is logarithmic. However,
the key point is that, as $p$ increases, the unstable behaviour
increases, regardless of parameter choice, which is corroborated by
both theory and experiments. This phenomenon was referred to in
\cite{Sanders2018}
%by Sanders et. al
as `the prevalence of chaotic dynamics in games with many
players'.
